---
title: "Data mining exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Introduction

Your collaborator has made a  CAGE experiment using 7 different tissues. As a remainder, CAGE tags are 20-21 nt long tags mapping to the genome. We often cluster CAGE tags that are close to one another on the genome to a “tag cluster”. A tag cluster can then have several tags from one or several tissues. 

One can view these CAGE tag clusters as being “core promoters” in the sense that they are measuring the activity and location of a core promoter.

The data file `htbinf_cage_tpms` shows the CAGE tag clusters as rows, and tissues as columns. The cell values are the TPMs from the given tissue in the cluster.  There are three additional columns: the tag cluster ID, the location of the cluster in mm8 and the strand of the cluster. 
The collaborator now wants to know:

__1. How many types of  core promoters are there in terms of tissue expression patterns, and what expression patterns are these? __

There are 4 types because we have 4 plots

```{r}
library(tidyverse)
library(viridis)
cage_data <- read_tsv("htbinf_cage_tpms.txt")
cage_data
```

```{r}
cage_pca <- prcomp(cage_data[c(-1:-3)])
cage_pca$x %>%
  as.data.frame() %>%
  ggplot(aes(x=PC1, y=PC2)) + geom_point()
```

```{r}
cage_data[c(-1:-3)] %>%
dist(.) %>%
hclust(.)%>%
plot(.)
```

```{r}
cage_kmeans <- kmeans(cage_data[c(-1:-3)], 3)
cage_data$clusters <- cage_kmeans$cluster
cage_data
  
```

```{r}
summary(cage_pca)
```



```{r}
cage_pca$x %>%
  as.data.frame() %>%
  mutate(clusters = as.factor(cage_kmeans$cluster)) %>%
  ggplot(aes(x=PC1, y=PC2, col=clusters)) + geom_point()
```

```{r}
library(pheatmap)
pheatmap(cage_data[c(-1:-3, -11)])
```



__2. What tissues are similar to each other in terms of promoter usage? They would really like to have this as a picture and not just  "values"__

```{r}
library(pheatmap)
pheatmap(cage_data[c(-1:-3)], color = magma(10))
```


__3.  How many tissue-specific promoters are there, per tissue and just allover? It would be very helpful to calculate two "specificity scores": one for each tissue, and one summary score across all tissues for each promoter. __ 

```{r}
cage_data <- read_tsv("htbinf_cage_tpms.txt")
cage_data_numeric <- cage_data[4:10]
rowwise_median <- apply(cage_data_numeric, 1, function(x) median(x))
cage_diff <- cage_data_numeric - rowwise_median
cage_diff_bool <- cage_diff > 20
cage_diff_bool %>%
  as.data.frame() %>%
  mutate(Sum = apply(cage_diff_bool, 1, sum)) -> cage_diff_bool_sum
cage_diff$Sum <- cage_diff_bool_sum$Sum
cage_diff$tc_id <- cage_data$tc_id
cage_diff %>%
  mutate(Var = apply(cage_diff[c(-8,-9)], 1, function(x) var(x))) %>%
  filter(Sum == 1) %>%
  filter(cer == max(cer) | emb == max(emb) | liv == max(liv) | lun == max(lun) | mac == max(mac) | mac == max(mac) | som == max(som) | vis == max(vis)) %>%
  arrange(desc(Var))
```

```{r}
cage_data %>%
  filter(tc_id == "C1F14437FA" | tc_id == "C7F861C25B")
```


```{r, eval = F}
rowwise_mean <- apply(cage_data[c(-1:-3, -11)], 1, function(x) mean(x))
rowwise_median <- apply(cage_data[c(-1:-3, -11)], 1, function(x) median(x))
rowwise_var <- apply(cage_data[c(-1:-3, -11)], 1, function(x) var(x))
rowwise_min <- apply(cage_data[c(-1:-3, -11)], 1, function(x) min(x) +0.001)
cage_diff <- cage_data[c(-1:-3, -11)]-rowwise_median
cage_diff %>%
  mutate(Var = apply(cage_mean_diff, 1, function(x) var(x))) %>%
  mutate("tc_id" = cage_data$tc_id) %>%
  arrange(desc(Var)) -> cage_diff_var
cage_diff_numeric <- cage_diff_var[1:7]
cage_diff_numeric
threshold <- 20
cage_diff_boolean <- cage_diff > threshold #quantile(as.matrix(cage_numeric), 0.95)
cage_diff_boolean %>%
  as.data.frame() %>%
  mutate("tc_id" = cage_data$tc_id) %>%
  mutate(Sum = apply(cage_boolean, 1, sum)) %>%
  arrange(desc(Sum)) %>% #-> cage_diff_boolean_sum #some
  filter(Sum == 1)-> cage_diff_boolean_filtered
  #filter(Sum == 1) #-> filtered_cage_boolean
#filtered_cage_boolean <- filtered_cage_boolean[1:7]
sapply(cage_diff_boolean_filtered[1:7], sum)
sapply(cage_diff_boolean_sum[1:7], sum)
```

```{r, eval = F}
threshold <- 20
cage_diff_var %>%
  filter(tc_id %in% cage_diff_boolean_filtered$tc_id) %>%
  gather("Tissue", "ExpDev", -Var, -tc_id) %>%
  group_by(Tissue) %>%
  summarize(Max = max(ExpDev))
  #mutate(TisSpec = ExpDev > threshold) #%>%
  #group_by(Tissue) %>%
  #summarize(Max = max(ExpDev), sum(TisSpec))
```



>Pick a cutoff for variance, high variance rows are uniquely expressed in some tissues (e.g. 100)
Pick a cutoff for the expression level (e.g 20) and then calculate a boolean matrix depending on this
Count the booleans



```{r, eval =F}
cage_data[c(-2:-3, -11)] %>%
  #rowwise() %>%
  #mutate(Quantile = apply(cage_data[c(-1:-3, -11)], 1, function (x) quantile(x, 0.95))) -> cage_mean
  mutate(Quantile = apply(cage_data[c(-1:-3, -11)], 1, function (x) quantile(x, 0.95))) -> cage_mean
  
  #as.data.frame() %>%
  #filter_if(is.numeric, any_vars(. >25))
  

cage_numeric <- cage_data[c(-1:-3, -11)]
cage_boolean <- cage_numeric > quantile(as.matrix(cage_numeric), 0.95)
cage_boolean %>%
  as.data.frame() %>%
  mutate("tc_id" = cage_data$tc_id) %>%
  mutate(Sum = apply(cage_boolean, 1, sum)) %>%
  filter(Sum == 1) -> filtered_cage_boolean
filtered_cage_boolean <- filtered_cage_boolean[1:7]
sapply(filtered_cage_boolean, sum)
```

```{r, eval =F}
sapply(filtered_cage_boolean, sum)
#t.test(cage_numeric[1], cage_numeric[2])
```



__4. They want to have  genome browser examples of the most tissue-specific promoter for each tissue__

```{r, eval=F}
cage_diff_var %>%
  filter(tc_id %in% cage_diff_boolean_filtered$tc_id)
  #group_by(Tissue) %>%
  #summarize(Max = max(ExpressionDeviation))
```



__5 Lastly, they want a list with the 10 most tissue-specific promoters, taking all tissues into account.__
```{r, eval =F}
pheatmap(cage_data[c(-1:-3)], color = rainbow(10))

```



